\documentclass[10pt]{article}
\usepackage{times}
\renewcommand{\ttdefault}{cmtt}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{setspace}
\singlespacing

\title{Genetic Algorithm-Based Fairness Violation Discovery\\in Machine Learning Classifiers}
\author{Student Name}
\date{}

\begin{document}
\maketitle

% ===================================================================
\section{Introduction}
% ===================================================================

Machine learning models increasingly drive high-stakes decisions---from credit scoring
and hiring shortlists to criminal recidivism predictions.  When these models treat
individuals differently based on protected attributes such as gender or race, the
consequences can be severe: the ProPublica investigation into the COMPAS recidivism
tool~[1] revealed that Black defendants were roughly twice as likely as white defendants
to be misclassified as high risk, despite comparable re-offence rates.  Similar concerns
have been documented in automated hiring systems and lending algorithms~[2].

Detecting such discrimination is difficult because the input space of a classifier is
typically high-dimensional, and unfair behaviour may only surface in specific regions
that are hard to find by inspection alone.  Exhaustive testing is infeasible, so the
problem naturally lends itself to \emph{search-based} methods that can explore the input
space efficiently within a limited evaluation budget.

I chose the \textit{AI Model Fairness Testing} problem from the module's baseline list
because it sits at the intersection of search-based software engineering and social
responsibility.  The specific question I investigate is: \textbf{given the same number of
model evaluations, can a genetic algorithm (GA) discover more discriminatory inputs than
a simple random search?}  I test this across three benchmark datasets with multiple
fairness metrics, paired statistical tests, convergence analysis, and hyperparameter
sensitivity experiments to give a thorough answer.

% ===================================================================
\section{Related Work}
% ===================================================================

\subsection{Fairness definitions and metrics}
More than twenty distinct fairness definitions have been proposed for classification
systems~[3].  At the group level, \emph{demographic parity} requires that positive
prediction rates be equal across demographic groups, while \emph{equalized odds} further
conditions on the true label~[4].  At the individual level, fairness asks whether two
individuals differing only in a protected attribute receive the same prediction.
Verma and Rubin~[3] show, using the German Credit dataset, that these definitions can
conflict---satisfying one may violate another---which motivates reporting multiple
metrics rather than relying on a single one.

\subsection{Fairness testing tools}
Galhotra et~al.\ introduced \textit{Themis}~[5], a tool that generates random test
inputs and checks whether software exhibits group-level causal discrimination.  Themis
is effective but relies entirely on random sampling, which can be wasteful when
discriminatory inputs are sparse.  Udeshi et~al.\ proposed \textit{Aequitas}~[6], which
uses directed probabilistic search strategies to find individual discriminatory
instances.  Their experiments on six classifiers showed that up to 70\% of generated
inputs were discriminatory and that retraining with those inputs improved fairness by up
to 94\%.  Zhang et~al.~[7] took a white-box approach, exploiting gradient information
and clustering to generate discriminatory instances 25 times faster than Aequitas on
deep neural networks.  However, white-box methods require access to model internals,
which is not always available.

\subsection{Search-based software engineering}
The idea of reformulating software engineering tasks as metaheuristic search problems
was formalised by Harman and Jones~[8].  Within this framework, random search serves as
a natural baseline because it requires no tuning, but it explores the space uniformly
and does not learn from previous evaluations.  Genetic algorithms maintain a
\emph{population} of candidate solutions that evolves through selection, crossover, and
mutation, gradually focusing on high-fitness regions~[9].  McMinn~[10] surveys the
extensive use of GAs for test data generation and notes that their effectiveness depends
heavily on the fitness landscape and operator design.

Compared to Themis (pure random) and Aequitas (local directed search), the GA-based
approach I use sits between the two: it is a global population-based search that does
not need model gradients (unlike~[7]) but can exploit structure in the fitness landscape
(unlike~[5]).

% ===================================================================
\section{Problem Definition}
% ===================================================================

Let $f$ be a binary classifier and $T_s(x)$ the operator that flips the sensitive
attribute of input~$x$.  The \emph{individual discrimination score} is
\begin{equation}\label{eq:disc}
D(x) = \bigl|P(y{=}1 \mid x) - P(y{=}1 \mid T_s(x))\bigr|.
\end{equation}
A large $D(x)$ means the model's positive-class probability changes substantially when
only the sensitive attribute is flipped, indicating potential unfairness toward that
individual.  I also report two group-level metrics computed over all generated samples:
demographic parity difference and an equalized-odds proxy (Section~\ref{sec:setup}).

% ===================================================================
\section{Method}
% ===================================================================

\subsection{Genetic algorithm design}

The population at generation~$t$ is
$\mathcal{P}^{(t)} = \{x_1, \ldots, x_N\}$, and the objective is
$\max_x D(x)$---i.e.\ discover the inputs with the largest fairness gap.

\paragraph{Tournament selection ($k{=}3$).}
I pick three individuals at random and keep the one with the highest cached fitness
value.  This gives moderate selection pressure without requiring a full population sort.
I considered $k{=}2$ (less pressure, slower convergence) and $k{=}5$ (faster convergence
but higher risk of premature loss of diversity); $k{=}3$ is a standard middle ground
recommended in the literature~[9].

\paragraph{Uniform crossover.}
Each feature of the child is copied from one of the two parents with probability~0.5.
I chose uniform crossover over single-point crossover because the features of a
tabular record are largely independent---there is no natural ordering that would make
a single cut point meaningful.

\paragraph{Gaussian mutation.}
With probability equal to the mutation rate ($p_m = 0.1$ by default), a single randomly
chosen feature receives additive noise $\mathcal{N}(0,\,0.1)$.  Gaussian noise is a
natural choice for continuous features and the small standard deviation keeps the
perturbation local, acting as fine-tuning around an already-promising region.  All values
are clipped to the training-data feature ranges so that candidates remain realistic.

\paragraph{Equal-budget comparison.}
Every model evaluation---including the initial population---counts toward the budget~$B$.
Both GA and random search receive exactly $B$~evaluations per trial, making the
comparison fair.  Tournament selection uses cached fitness values rather than additional
model calls, so no hidden evaluations inflate the GA's budget.

\subsection{Random search baseline}
The baseline generates $B$~independent random samples from the same feature ranges and
evaluates each one.  It serves as a lower bound: any method that cannot outperform random
search is arguably not learning useful structure.

% ===================================================================
\section{Experiments}
% ===================================================================

\subsection{Setup}\label{sec:setup}
\begin{itemize}
    \item \textbf{Datasets}: Adult (48\,842 records, 14 features), COMPAS
          (7\,214 records, 8 features), German Credit (1\,000 records, 20 features).
    \item \textbf{Sensitive attributes}: \texttt{sex} (Adult),
          \texttt{race} (COMPAS, binarised as African-American~$=0$ vs.\ all
          other races~$=1$), \texttt{sex} (German).
    \item \textbf{Classifier}: \texttt{RandomForestClassifier} (scikit-learn,
          default hyperparameters, \texttt{random\_state=42}).
    \item \textbf{Trials}: 30 independent runs with different random seeds.
    \item \textbf{Budget}: $B = 50\,000$ evaluations per method per trial.
    \item \textbf{GA defaults}: population size $N{=}50$, mutation rate
          $p_m{=}0.1$, tournament size $k{=}3$.
    \item \textbf{Baseline}: Random search with the same budget.
    \item \textbf{Metrics} (recorded per trial):
    \begin{enumerate}
        \item Individual discrimination count:
              $|\{x : D(x) > 0.2\}|$.
        \item Mean individual discrimination score:
              $\bar{D}$.
        \item Demographic parity difference (DP diff):
              $|\text{rate}_{g=0} - \text{rate}_{g=1}|$ computed over
              \emph{generated} (not real-data) samples.  Because GA-generated
              samples are biased toward discriminatory regions, this metric
              reflects search effectiveness rather than the model's overall
              demographic parity.
        \item Counterfactual prediction gap (EO proxy):
              the maximum group-rate difference conditioned on counterfactual
              model predictions obtained by flipping the sensitive attribute.
              This is \emph{not} standard equalized odds, which conditions on
              true labels; generated inputs have no ground-truth labels, so
              counterfactual predictions serve as a proxy.
    \end{enumerate}
\end{itemize}

\subsection{Procedure}
For every dataset the following pipeline runs:
(1)~load and preprocess data, encoding categorical features and normalising the
sensitive attribute to binary;
(2)~train the classifier on a 70/30 stratified split;
(3)~for each of the 30 trials, execute GA and random search with the same seed and
budget, recording all four metrics and the per-generation convergence trace;
(4)~optionally, run a hyperparameter sensitivity sweep over population size
$\in \{10, 20, 40\}$ and mutation rate $\in \{0.05, 0.1, 0.2\}$, keeping the budget
constant.

\subsection{Statistical test}
I use the one-sided Wilcoxon signed-rank test~[11] at $\alpha{=}0.05$ with alternative
hypothesis GA~$>$~Random for each dataset--metric pair.  With 30~paired observations per
test, the minimum attainable $p$-value is approximately $9.3 \times 10^{-10}$, providing
ample power to detect real differences.

% ===================================================================
\section{Results}
% ===================================================================

\subsection{Overall comparison}

Table~\ref{tab:main-results} summarises the 30-trial results.  GA outperforms random
search on \textbf{11 out of 12} dataset--metric combinations at $\alpha{=}0.05$.

\begin{table}[h]
    \centering
    \caption{GA vs Random Search (30 trials, $B{=}50\,000$).  $p$-values from one-sided
    Wilcoxon signed-rank test.}
    \label{tab:main-results}
    \begin{tabular}{llccc}
        \toprule
        Dataset & Metric & GA (mean$\pm$std) & Random (mean$\pm$std) & $p$ \\
        \midrule
        Adult  & ID count & $49348 \pm 345$   & $4.5 \pm 1.8$       & ${<}\,10^{-6}$ \\
        Adult  & ID mean  & $0.239 \pm 0.028$ & $0.041 \pm 0.000$   & ${<}\,10^{-9}$ \\
        Adult  & DP diff  & $0.252 \pm 0.375$ & $0.006 \pm 0.002$   & $0.008$ \\
        Adult  & EO diff  & $0.500 \pm 0.392$ & $0.081 \pm 0.009$   & ${<}\,10^{-5}$ \\
        \midrule
        COMPAS & ID count & $49447 \pm 941$   & $269 \pm 17$        & ${<}\,10^{-6}$ \\
        COMPAS & ID mean  & $0.365 \pm 0.104$ & $0.052 \pm 0.000$   & ${<}\,10^{-9}$ \\
        COMPAS & DP diff  & $0.428 \pm 0.466$ & $0.002 \pm 0.001$   & $0.004$ \\
        COMPAS & EO diff  & $0.567 \pm 0.420$ & $0.034 \pm 0.021$   & ${<}\,10^{-6}$ \\
        \midrule
        German & ID count & $3257 \pm 12188$  & $0.0 \pm 0.0$       & $0.090$ \\
        German & ID mean  & $0.155 \pm 0.022$ & $0.027 \pm 0.000$   & ${<}\,10^{-9}$ \\
        German & DP diff  & $0.367 \pm 0.397$ & $0.038 \pm 0.004$   & ${<}\,10^{-4}$ \\
        German & EO diff  & $0.693 \pm 0.269$ & $0.152 \pm 0.005$   & ${<}\,10^{-9}$ \\
        \bottomrule
    \end{tabular}
\end{table}

\subsection{Convergence analysis}
The convergence plots (generated by \texttt{visualization.py}) show the expected
pattern: the GA's best-per-generation discrimination score rises steeply in the first
few hundred generations and then levels off, while random search's chunk-best stays
approximately flat throughout the budget.  This confirms that the GA is actively
\emph{learning} where discriminatory regions lie rather than finding them by luck.

On COMPAS the convergence is fastest---the GA saturates to near-maximum fitness within
roughly 200 generations---which is consistent with the model being highly sensitive to
the \texttt{race} attribute.  On German the curve is lower overall, reflecting the
weaker dependence on \texttt{sex}.

\subsection{Hyperparameter sensitivity}
The sensitivity sweep (population size $\in \{10,20,40\}$, mutation rate
$\in \{0.05, 0.1, 0.2\}$, same total budget) shows two trends:
\begin{itemize}
    \item Larger populations consistently improve ID-count discovery.  At $N{=}40$
          the GA finds ${\sim}\,49\,500$ discriminatory inputs on Adult, compared to
          near-zero at $N{=}10$ with $p_m{=}0.05$.  A larger population provides more
          diversity for selection and crossover to exploit.
    \item The effect of mutation rate is smaller but still noticeable: on Adult,
          $p_m{=}0.1$ and $p_m{=}0.2$ with $N{=}40$ both achieve ${\sim}\,49\,500$
          ID-count, while $p_m{=}0.05$ drops slightly.  Moderate mutation keeps
          exploration active without disrupting converged solutions.
\end{itemize}
The default configuration ($N{=}50$, $p_m{=}0.1$) falls in the high-performance region
of this grid, supporting the parameter choices.

\subsection{Discussion}

The most striking result is the sheer scale of the gap: on Adult and COMPAS, the GA
discovers roughly \emph{all} of its 50\,000 evaluated inputs as discriminatory
($D(x)>0.2$), whereas random search finds only a handful.  This happens because the GA
quickly converges to input-space regions where the classifier's predictions flip when the
sensitive attribute changes.  Random search, by contrast, samples uniformly across the
entire feature space, most of which has low discrimination.

\textbf{German Credit is the exception.}  The individual-count metric is the only one of
the twelve that is \emph{not} statistically significant ($p{=}0.09$).  In 28 of the 30
trials, neither method finds any input exceeding the 0.2 threshold.  Two GA trials do
find ${\sim}\,49\,000$ such inputs, but the remaining 28~zeros dominate the test.  The
other three metrics (ID mean, DP diff, EO diff) are all highly significant even on
German, showing that the GA does find larger discrimination---just not large enough to
clear 0.2 consistently.  This suggests the fixed threshold is too aggressive for this
dataset.

The high variance of demographic parity for GA (e.g.\ $0.252 \pm 0.375$ on Adult)
deserves attention.  Because the GA directs search toward high-discrimination regions, the
generated samples are not representative of the overall input distribution, so the
group-level rate difference fluctuates heavily between trials.  This is not a failure of
the GA; it is an artefact of measuring a group-level metric on a non-uniform sample.
Readers should therefore interpret the DP and EO-proxy columns in Table~\ref{tab:main-results}
as measures of \emph{search-biased fairness gap}---i.e.\ how large a disparity the search
method can uncover---rather than as estimates of the model's demographic parity or
equalized odds on the real data distribution.

% ===================================================================
\section{Reflection}
% ===================================================================

There are several limitations to acknowledge:
\begin{itemize}
    \item \textbf{Generated vs.\ real inputs.} Fairness metrics are computed over
          synthetic samples drawn uniformly from training-data feature ranges, not from
          the real data distribution.  Discrimination found in low-density regions may
          rarely occur in practice.
    \item \textbf{Counterfactual prediction gap (EO proxy).} Because generated inputs
          have no ground-truth labels, I condition on counterfactual model predictions
          (flipping the sensitive attribute) rather than true labels.  This metric is
          therefore \emph{not} standard equalized odds; it is closer to a counterfactual
          fairness measure and may over- or under-estimate the true equalized-odds gap.
    \item \textbf{Fixed threshold.} The $D(x) > 0.2$ threshold works well for Adult and
          COMPAS but is too aggressive for German Credit.  An adaptive or
          dataset-specific threshold would give a more nuanced view.
    \item \textbf{Single classifier.} All experiments use RandomForest.  A logistic
          regression or neural network might exhibit different sensitivity patterns,
          which could change the relative advantage of GA over random search.
    \item \textbf{GA hyperparameters.} The sensitivity analysis covers a $3{\times}3$
          grid, which is not exhaustive.  More advanced tuning (e.g.\ adaptive mutation
          or self-adaptive parameter control) could further improve performance.
    \item \textbf{Scalability.} The current implementation evaluates one sample at a
          time.  Batch prediction would substantially reduce wall-clock time and make
          larger budgets feasible.
\end{itemize}

Future directions include multi-objective optimisation (jointly maximising discrimination
discovery and input realism), constraint-aware sampling to keep generated inputs closer
to the real data manifold, batch evaluation for scalability, and testing across multiple
classifier types to check whether the GA's advantage generalises.

% ===================================================================
\section{Conclusion}
% ===================================================================

This project presents a reproducible, budget-controlled pipeline for comparing
GA-based fairness testing against random search.  Across three benchmark datasets, four
fairness metrics, 30 independent trials, paired Wilcoxon signed-rank tests, convergence
analysis, and hyperparameter sensitivity experiments, the GA \textbf{significantly
outperforms random search in 11 out of 12 comparisons} ($p < 0.05$).

On Adult and COMPAS the improvement is dramatic: the GA discovers tens of thousands of
discriminatory inputs per trial while random search finds fewer than 300.  On German
Credit the advantage is present but weaker due to the model's lower sensitivity to the
protected attribute and the fixed discrimination threshold.

The convergence analysis confirms that the GA genuinely \emph{learns} where
discriminatory inputs concentrate, and the sensitivity analysis shows that performance
is robust across a range of population sizes and mutation rates---provided the
population is not too small.  These findings support the practical value of
evolutionary search for fairness testing under limited evaluation budgets.

% ===================================================================
\section{Artifact}
% ===================================================================

Code repository: \texttt{<REPLACE-WITH-PUBLIC-REPO-URL>}\\
Raw results: \texttt{experiments/results/}\\
Replication:
\begin{enumerate}
    \item \texttt{pip install -r requirements.txt}
    \item \texttt{python scripts/download\_adult.py}
    \item \texttt{python scripts/download\_benchmark\_datasets.py}
    \item \texttt{python experiments/run\_experiment.py --trials 30 --budget 50000
          --with-sensitivity}
    \item \texttt{python experiments/statistical\_test.py}
    \item \texttt{python experiments/visualization.py --dataset adult}
\end{enumerate}

% ===================================================================
\section{References}
% ===================================================================

\begin{small}
[1] J.~Angwin, J.~Larson, S.~Mattu, and L.~Kirchner, ``Machine Bias,''
\textit{ProPublica}, May~2016.

[2] S.~Barocas and A.~D.~Selbst, ``Big Data's Disparate Impact,''
\textit{California Law Review}, vol.~104, no.~3, pp.~671--732, 2016.

[3] S.~Verma and J.~Rubin, ``Fairness Definitions Explained,'' in
\textit{Proc.\ FairWare}, 2018, pp.~1--7.

[4] M.~Hardt, E.~Price, and N.~Srebro, ``Equality of Opportunity in Supervised
Learning,'' in \textit{Proc.\ NeurIPS}, 2016, pp.~3315--3323.

[5] S.~Galhotra, Y.~Brun, and A.~Meliou, ``Fairness Testing: Testing Software for
Discrimination,'' in \textit{Proc.\ ESEC/FSE}, 2017, pp.~498--510.

[6] S.~Udeshi, P.~Arora, and S.~Chattopadhyay, ``Automated Directed Fairness
Testing,'' in \textit{Proc.\ ASE}, 2018, pp.~98--108.

[7] P.~Zhang, J.~Wang, J.~Sun, et~al., ``White-box Fairness Testing through
Adversarial Sampling,'' in \textit{Proc.\ ICSE}, 2020, pp.~949--960.

[8] M.~Harman and B.~F.~Jones, ``Search-based software engineering,''
\textit{Information and Software Technology}, vol.~43, no.~14, pp.~833--839, 2001.

[9] J.~H.~Holland, \textit{Adaptation in Natural and Artificial Systems},
University of Michigan Press, 1975.

[10] P.~McMinn, ``Search-based software test data generation: a survey,''
\textit{Software Testing, Verification and Reliability}, vol.~14, no.~2,
pp.~105--156, 2004.

[11] F.~Wilcoxon, ``Individual comparisons by ranking methods,''
\textit{Biometrics Bulletin}, vol.~1, no.~6, pp.~80--83, 1945.
\end{small}

\end{document}
